---
title: "Differential expression of genes and modules"
subtitle: "J. Shah chimeric mouse collaboration"
author: "Kim Dill-McFarland, kadm@uw.edu"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
date: "version `r format(Sys.time(), '%B %d, %Y')`"
editor_options: 
  chunk_output_type: console
---

# Background

The purpose of this workflow is to identify differentially expressed (DE) genes and modules.

# Setup
Load packages

```{r message=FALSE, warning=FALSE}
# Data manipulation and figures
library(tidyverse)
    # Multi-panel figures for ggplot
    library(cowplot)

#Define ggplot colors
logFC.cols <- c("Down, FDR < 0.5"="lightblue",
                "Down, FDR < 0.2"="blue",
                "Down, FDR < 0.05"="darkblue",
                "NS"="grey",
                "Up, FDR < 0.5"="pink",
                "Up, FDR < 0.2"="red",
                "Up, FDR < 0.05"="darkred")

#Linear models
library(limma)
#Construct networks to ID modules
library(WGCNA)
# Print tty table to knit file
library(knitr)
library(kableExtra)
options(knitr.kable.NA = '')
```

Set seed

```{r}
set.seed(4389)
```

Scripts

```{r}
source("https://raw.githubusercontent.com/kdillmcfarland/R_bioinformatic_scripts/master/RNAseq_module_fxn.R")
source("https://raw.githubusercontent.com/kdillmcfarland/R_bioinformatic_scripts/master/limma.extract.pval.R")
source("scripts/RNAseq_boxplot_fxn.R")

`%notin%` <- Negate(`%in%`)
```

Set variable names and cutoffs for this workflow.

```{r}
#Rdata file WITHIN project directory that holds cleaned data
data.file <- "data_clean/Shah.clean.RData"

#Prefix to give file names
basename <- "Shah_contrast"
#Define variable(s) of interest
#Used in PCA plots and to select significant genes to be used in module building
vars_of_interest <- c("status","cell")

#Maximum fdr for genes to be included in plots and modules
gene.fdr.cutoff <- 0.5
```

# Load data

```{r}
#Load data
load(data.file)
```

This includes in the following samples.

```{r echo=FALSE}
dat.voom$targets %>% 
  group_by_at(vars(all_of(vars_of_interest))) %>% 
  tally() %>% 
  
  kable(align="c") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

# Data exploration
## PCA (genes)

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=6}
#Calculate PCA for all data.
PCA.all <- as.data.frame(dat.voom$E) %>% 
  t() %>% 
  #Calc PCA
  prcomp()

PC1.label <- paste("PC1 (", summary(PCA.all)$importance[2,1]*100, "%)", sep="")
PC2.label <-paste("PC2 (", summary(PCA.all)$importance[2,2]*100, "%)", sep="")

# Extract PC values
PCA.all.dat <- as.data.frame(PCA.all$x) %>% 
  rownames_to_column("libID") %>%
  # Merge with metadata
  full_join(as.data.frame(dat.voom$targets), by="libID")

PCA1 <- ggplot(PCA.all.dat,
                  aes(PC1, PC2)) +
           geom_point(aes(color=status),
                      size=3) + 
  #Beautify
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  labs(x=PC1.label, y=PC2.label, title="voom normalized logCPM") +
  coord_fixed(ratio=1)

PCA2 <- ggplot(PCA.all.dat,
                  aes(PC1, PC2)) +
           geom_point(aes(color=cell),
                      size=3) + 
  #Beautify
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  labs(x=PC1.label, y=PC2.label, title="voom normalized logCPM") +
  coord_fixed(ratio=1) +
  scale_color_brewer(palette="Set1")

PCA3 <- ggplot(PCA.all.dat,
                  aes(PC1, PC2)) +
           geom_point(aes(color=status, shape=cell),
                      size=3) + 
  #Beautify
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  labs(x=PC1.label, y=PC2.label, title="voom normalized logCPM") +
  coord_fixed(ratio=1)

PCA4 <- ggplot(PCA.all.dat,
                  aes(PC1, PC2)) +
           geom_point(aes(color=as.character(mouse)),
                      size=3) + 
  #Beautify
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  labs(x=PC1.label, y=PC2.label, title="voom normalized logCPM") +
  coord_fixed(ratio=1)+
  scale_color_brewer(palette="Set3")

plot_grid(PCA1, PCA2, PCA3, PCA4, ncol=2)

ggsave("figs/PCA_Shah_geneVoom.png", 
       plot_grid(PCA1, PCA2, PCA3, PCA4, align = "hv",
                 ncol=2),
       height=8, width=10)
```

# Define significant genes
## Linear model 

```{r}
# Define model
model <- model.matrix(~0 + status:cell, data=dat.voom$targets)
    colnames(model) <- c("Uninfected_WT", "Infected_WT",
                         "Uninfected_TKO", "Infected_TKO")
# Define contrasts of interest
contrasts <- makeContrasts(
  uninfected = Uninfected_TKO-Uninfected_WT, 
  infected = Infected_TKO-Infected_WT,
  levels=model)
  
#Calculate consensus corr of samples from the same donor
consensus.corr <- duplicateCorrelation(
                    dat.voom$E,
                    model,
  block=dat.voom$targets$mouse)$consensus.correlation
  
  consensus.corr
  
# Fit model to transformed count data. Calculate eBayes
  efitQW <- eBayes(
            contrasts.fit(contrasts=contrasts,
            lmFit(dat.voom$E, model,
                  block=dat.voom$targets$mouse,
                  correlation=consensus.corr)))
```

```{r warning=FALSE, message=FALSE, echo=FALSE}
gene.result.name <- paste(basename, "gene_pval", sep="_")
#Extract p-values from results
extract.pval(model=model,
             voom.dat=dat.voom$E, 
             eFit=efitQW, 
             name=gene.result.name,
             summary=TRUE,
             contrasts=TRUE, contrast.mat=contrasts)

#Write to disk
dir.create(path="results/gene_level/", showWarnings = FALSE)
get(gene.result.name) %>% 
  #Add annotations
  left_join(dat.voom$genes, by="geneName") %>% 
  dplyr::select(geneName, mgi_symbol, gene_biotype,
                everything()) %>% 
#Save
write_csv(paste("results/gene_level/", gene.result.name,
                ".csv", sep=""))
```

### Summarize gene model

```{r echo=FALSE}
get(paste(gene.result.name, "summ", sep=".")) %>% 
  mutate(comparison = "TKO-WT") %>% 
  select(group, comparison, everything()) %>% 
  
  kable(align=c("l","c","c","c","c","c","c"),
        col.names = c("Variable", "Comparison", "0.05", "0.1", "0.2","0.3","0.4","0.5")) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>% 
  add_header_above(c(" "=2, "Genes with FDR <"=6))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, fig.width=9}
get(gene.result.name) %>%  
  filter(group != "(Intercept)") %>% 
  mutate(col.group = ifelse(adj.P.Val <= 0.05 & FC.group=="up", 
                            "Up, FDR < 0.05",
                     ifelse(adj.P.Val <= 0.05 & FC.group=="down", 
                            "Down, FDR < 0.05",
                     ifelse(adj.P.Val <= 0.2 & FC.group=="up", 
                            "Up, FDR < 0.2",
                     ifelse(adj.P.Val <= 0.2 & FC.group=="down", 
                            "Down, FDR < 0.2",
                     ifelse(adj.P.Val <= 0.5 & FC.group=="up", 
                            "Up, FDR < 0.5",
                     ifelse(adj.P.Val <= 0.5 & FC.group=="down", 
                            "Down, FDR < 0.5",
                            "NS"))))))) %>%
  arrange(group,-adj.P.Val) %>% 
  
ggplot(aes(x=AveExpr, y=logFC, color=col.group)) +
  geom_point(size=2) +
  scale_color_manual(values=logFC.cols) +
  facet_grid(~group, scales = "free_y")+
  theme_classic() +
  labs(x="Average log CPM", y="Log fold change", color="") +
  guides(color = guide_legend(reverse = TRUE)) +
  theme(text = element_text(size=18),
        legend.position = "bottom") +
  guides(color=guide_legend(nrow=3, byrow=TRUE))
```

### Compare significant genes

```{r echo=FALSE, message=FALSE}
Shah_gene_pval <- read_csv("results/gene_level/Shah_gene_pval.csv")

par(mfrow = c(2, 3))
library(venn)
overlap.summ <- data.frame()

for(fdr in c(0.05,0.1,0.2,0.3,0.4,0.5)){
  #filter signif genes
  signif1 <- Shah_gene_pval %>% 
    filter(group != "(Intercept)" & adj.P.Val <= fdr) %>% 
    select(geneName) %>% 
    distinct() %>% unlist(use.names = FALSE)
  
  signif2 <- Shah_contrast_gene_pval %>% 
    filter(adj.P.Val <= fdr) %>% 
    select(geneName) %>% 
    distinct() %>% unlist(use.names = FALSE)
  
  #Venn
  venn(ilab=FALSE, zcolor = "style",ilcs=1.2, sncs=1.2,
     snames=c("interaction_model","contrast_model"),
     x=list(signif1,signif2))
  
  title(sub=paste("FDR <", fdr, sep=" "),line = -1, cex.sub=1.5)
  
  #Summary table
  summ1 <- Shah_gene_pval %>% 
    filter(group != "(Intercept)" & adj.P.Val <= fdr & 
             geneName %notin% signif2) %>% 
    group_by(geneName) %>% 
    summarize(signif = paste(group, collapse=",")) %>% 
    ungroup() %>% 
    count(signif)
  
  summ2 <- Shah_contrast_gene_pval %>% 
    filter(group != "(Intercept)" & adj.P.Val <= fdr & 
             geneName %notin% signif1) %>% 
    group_by(geneName) %>% 
    summarize(signif = paste(group, collapse=",")) %>% 
    ungroup() %>% 
    count(signif)
  
  summ.all <- bind_rows(summ1, summ2) %>% 
    mutate(model = c("interaction","contrast"),
           FDR = fdr)
  overlap.summ <- bind_rows(overlap.summ, summ.all)
}

overlap.summ %>% 
  select(FDR, model, n, signif) %>% 
  
  kable(align="c", "html",
        col.names = c("FDR", "model", 
          "genes not signif in other model",
          "variable(s) those gene are signif for in this model")) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>% 
  row_spec(c(2,4,6,8,10), extra_css = "border-bottom: 1px solid")
```

Roughly half of genes identified as significant in either model are also significant in the other model. Genes unique to the interaction model are only significant for infection `status`. Thus, the contrasts model appears to be successfully removing much of the infection-only signal while retaining all `cell` type and `status:cell` interaction signals.

Interestingly the genes unique to the contrasts model were only significant for cell type within `infected` cells. This may indicate that the contrasts model is better resolving signal from the interaction term `status:cell` since the interaction model defines the interaction from the intercept of `status` = uninfected and `cell` = WT.

Overall, these results indicate that the contrasts model better resolves the signals of interest.

## Gene plots

Save in `figs/gene_level_contrast`

```{r eval=FALSE, echo=FALSE}
#Subset to genes for modules 
genes.toPlot <- unique(Shah_contrast_gene_pval$geneName)

#PLOTS
dir.create("figs/gene_level_contrast/", showWarnings = FALSE)
plot.all(voom.dat=dat.voom, 
         pval.dat=get(gene.result.name), 
         genes.toPlot=genes.toPlot,
         gene.key="data_clean/ensembl.key.txt",
         vars=c("status","cell"),
         interaction=TRUE,
         interaction.terms=c("uninfected","infected"),
         color.var="mouse",
         outdir="figs/gene_level_contrast/", 
         name=paste(basename,"expression_", sep="_"),
         cores=3, width=5, height=7)
```

# Modules: Status and cell

Define customizations for module building.

```{r}
#Set FDR cutoff for gene inclusion in modules
mod.fdr.cutoff <- 0.3
#List variables from which significant genes will be extracted
vars_for_mods <- c("uninfected","infected")
```

```{r echo=FALSE}
genes.signif <- get(gene.result.name) %>% 
  filter(group %in% vars_for_mods) %>% 
  filter(adj.P.Val <= mod.fdr.cutoff) %>% 
  distinct(geneName) %>% 
  select(geneName) %>% unlist(use.names = FALSE)
```

In total, `r length(genes.signif)` of `r nrow(dat.voom$E)` genes that significantly differed (FDR $\leq$ `r mod.fdr.cutoff`) by one or more variables of interest will be incorporated into gene modules.

```{r}
make.modules(voom.dat = dat.voom,
             genes.signif = genes.signif,
             Rsq.min = 0.8,
             minModuleSize = 50,
             deepSplit = 3,
             nThread = 4,
             basename = basename)
```

```{r message=FALSE, echo=FALSE}
#Load results
mod.result.dir <- list.files(path = "results",
                             pattern = basename,
                             full.names = TRUE)

mod.result.file <- list.files(path = mod.result.dir,
                            pattern = "genes_in_mod.csv",
                            full.names = TRUE)

mods.net <- read_csv(mod.result.file) %>% 
  #add annotation
  select(-mgi_symbol, -gene_biotype) %>% 
  left_join(dat.voom$genes, by="geneName") %>% 
  select(geneName, module, mgi_symbol, gene_biotype,
         module.char, mod.color)
  
  write_csv(mods.net, mod.result.file)
  
voom.result.file <- list.files(path = mod.result.dir,
                            pattern = "mod_voom_counts.csv",
                            full.names = TRUE)
voom.mods <- read_csv(voom.result.file) %>% 
  column_to_rownames("module")
```

A power threshold of `r sft.select$Power` was used as it achieves high R^2^ and sufficient mean connectivity (R^2^ = `r sft.select$SFT.R.sq`, mean k = `r sft.select$mean.k`).

```{r echo=FALSE}
#Number of genes in each module
kable(as.data.frame(table(mods.net$module.char)), 
      align="c", col.names = c("Module","Total genes")) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

This created `r max(mods.net$module)` modules plus `r nrow(mods.net[mods.net$module ==0,])` (`r nrow(mods.net[mods.net$module ==0,]) / length(genes.signif) * 100`%) genes not grouped into any module (*e.g.* in module 0).

## PCA (modules)

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=6}
# Calculate PCA
PCA2 <-  voom.mods %>% 
  t() %>% 
  prcomp()

PC1.label <- paste("PC1 (", summary(PCA2)$importance[2,1]*100, "%)", sep="")
PC2.label <-paste("PC2 (", summary(PCA2)$importance[2,2]*100, "%)", sep="")

# Extract PC values
PCA.all.dat <- as.data.frame(PCA2$x) %>% 
  rownames_to_column("libID") %>%
  # Merge with metadata
  full_join(as.data.frame(dat.voom$targets), by="libID")

PCA1 <- ggplot(PCA.all.dat,
                  aes(PC1, PC2)) +
           geom_point(aes(color=status),
                      size=3) + 
  #Beautify
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  labs(x=PC1.label, y=PC2.label, title="modules\nvoom normalized logCPM") +
  coord_fixed(ratio=1)

PCA2 <- ggplot(PCA.all.dat,
                  aes(PC1, PC2)) +
           geom_point(aes(color=cell),
                      size=3) + 
  #Beautify
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  labs(x=PC1.label, y=PC2.label, title="modules\nvoom normalized logCPM") +
  coord_fixed(ratio=1) +
  scale_color_brewer(palette="Set1")

PCA3 <- ggplot(PCA.all.dat,
                  aes(PC1, PC2)) +
           geom_point(aes(color=status, shape=cell),
                      size=3) + 
  #Beautify
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  labs(x=PC1.label, y=PC2.label, title="modules\nvoom normalized logCPM") +
  coord_fixed(ratio=1)

PCA4 <- ggplot(PCA.all.dat,
                  aes(PC1, PC2)) +
           geom_point(aes(color=as.character(mouse)),
                      size=3) + 
  #Beautify
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  labs(x=PC1.label, y=PC2.label, title="modules\nvoom normalized logCPM") +
  coord_fixed(ratio=1)+
  scale_color_brewer(palette="Set3")

plot_grid(PCA1, PCA2, PCA3, PCA4, ncol=2)

ggsave("figs/PCA_Shah_contrast_moduleVoom.png", 
       plot_grid(PCA1, PCA2, PCA3, PCA4, align = "hv",
                 ncol=2),
       height=8, width=10)
```

## Linear model

```{r}
#Calculate consensus corr of samples from the same donor   
consensus.corr <- duplicateCorrelation(
                    voom.mods,
                    model,
  block=dat.voom$targets$mouse)$consensus.correlation
  
  consensus.corr
  
# Fit model to transformed count data. Calculate eBayes
  efitQW.mods <- eBayes(
            contrasts.fit(contrasts=contrasts,
            lmFit(voom.mods, model,
                  block=dat.voom$targets$mouse,
                  correlation=consensus.corr)))
```

```{r warning=FALSE, message=FALSE, echo=FALSE}
mod.result.name <- paste(basename, "mod_pval", sep="_")

#Extract p-values from results
extract.pval(model=model,
             voom.dat=voom.mods, 
             eFit=efitQW.mods, 
             name=mod.result.name,
             summary=TRUE,
             contrasts=TRUE, contrast.mat=contrasts)

#Save to disk
get(mod.result.name) %>% 
  rename(module=geneName) %>% 
write_csv(.,
  paste(mod.result.dir, "/", mod.result.name,
                ".csv", sep=""))
```

### Summarize module model

```{r echo=FALSE}
get(paste(mod.result.name, "summ", sep=".")) %>% 
  mutate(comparison = "TKO-WT") %>% 
  select(group, comparison, everything()) %>% 
  
  kable(align=c("l","c","c","c","c","c","c"),
    col.names = c("Variable", "Comparison",
                  "0.05","0.1","0.2","0.3","0.4","0.5")) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>% 
  add_header_above(c(" "=2, "Modules with FDR <"=6))
```

```{r echo=FALSE}
get(mod.result.name) %>% 
  mutate(signif.05 = ifelse(adj.P.Val <= 0.05, 
                            as.character(FC.group), 
                            NA)) %>% 
  select(geneName, signif.05, group) %>% 
  pivot_wider(names_from = group, values_from = signif.05) %>% 
  mutate(force.group = ifelse(is.na(uninfected) | is.na(infected), 
                              "A","B")) %>% 
  arrange(force.group,uninfected,infected) %>% 
  select(-force.group) %>% 
  
  kable(align="c",
        col.names = c("module", 
                      "uninfected \ndirection in TKO", 
                      "infected \ndirection in TKO")) %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE) %>% 
  pack_rows("Significant in uninfected", 1,3) %>% 
  pack_rows("Significant in infected", 4,9) %>% 
  pack_rows("Significant in both", 10,19)
```

_Directions are relative to WT. For example, `down` means the module has **lower** expression in TKO relative to WT._

## Module plots

Create expression plots of modules, and save in `figs/module*`

```{r message=FALSE, echo=FALSE, results=FALSE, warning=FALSE}
mod.figs.dir <- list.files(path = "figs",
                             pattern = basename,
                             full.names = TRUE)
#Remove plots if exist
do.call(file.remove, list(list.files(mod.figs.dir,
                     pattern="module_[0-9]{1,4}.pdf",
                     full.names = TRUE)))

#PLOTS
plot.all(voom.dat=voom.mods, 
         pval.dat=get(mod.result.name), 
         meta.dat=as.data.frame(dat.voom$targets),
         genes.toPlot=unique(rownames(voom.mods))[-1],
         ######
         vars=c("status","cell"),
         interaction=TRUE,
         interaction.terms=c("uninfected","infected"),
         color.var="mouse",
         outdir=paste(mod.figs.dir[1],"/",sep=""), 
         name=paste(basename,"expression_mod_", sep="_"),
         cores=3, width=5, height=7)
```

# R session

```{r}
sessionInfo()
```

***